RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data

### 2013-11-21 doc/bioinformatics2013/RVD2.tex

BAM files for the synthetic DNA data were down sampled by $10\times$, $100\times$, $1,000\times$, $10,000\times$ to simulate lower coverage data while retaining the error structure of real NGS data. The final data set contains pairs of reads for three replicates of each case and a pair of reads for the control sample giving a total of 36 BAM files for each downsampled data set.

@@

GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples. Therefore, we expect those algorithms to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples. \citet{Stead:2013fu} showed that varscan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000. We find its performance on much lower MAF variants and across a wider range of coverage depths. Varscan2-somatic, streak and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the median overall median depth of the replicates.

>>>>

BAM files for the synthetic DNA data were down sampled by $10\times$, $100\times$, $1,000\times$, $10,000\times$ using Picard Version 1.96 to simulate lower coverage data while retaining the error structure of real NGS data. The final data set contains pairs of reads for three replicates of each case and a pair of reads for the control sample giving a total of 36 BAM files for each downsampled data set.

@@

\subsubsection*{Parameters setting for variant calling}\

The variant calling algorithms we compare RVD2 to include samtools/bcftools, GATK VarScan2-mpileup, Varscan2-somatic, Strelka, and muTect.

\paragraph{\textbf{Samtools}}We used samtools/bcftools (v0.1.19) for variant calling. mpileup was the function to call variants and bcftools was used to save the result in standard Variant Call Format (VCF) files. In mpileup, we set the -d option sufficiently high at $10^6$ to avoid coverage cap. Our synthetic data was mapped using BWA Version 0.7.4, -C50 option was applied to minimize the effect from reads with overestimated mapping quality. mpileup was able to accept multiple replicates at one time so six bam files from each replicate group were fed into mpileup for variant calling. Read depth for each sample was kept with -D enabled, and -u was turned on to make sure the output bcf files uncompressed.

\paragraph{\textbf{GATK}}
UnifiedGenotyper is the function we used in GATK (v2.1-8) to detect mutations.We followed the general GATK recomended workflow to detect variants. Due to some format  incompatibility, we applied Picard to format read group and GATK for realignment as preprocessing. When calling variants using UnifiedGenotyper, -ploid (Number of samples in each pool $\times$ Sample Ploidy) was set at 1 because our synthetic data is haploid; we change the default setting for -dcov (default max coverage cutoff of $250\times$ ) to $10^6$ to avoid downsamplig so that the coverage was comparable to other approaches. All other parameters were set at default. Six replicates was passed to GATK for variant calling.

\paragraph{\textbf{VarScan2-mpileup}}
VarScan2 (v2.3.4) mpileup2snp is a SNP calling program which takes multi-samples from samtools mpileup pipeline. We set -d at $10^6$ and enabled -50 for same concern previously stated. In afterwards mpileup2snp function, --min-var-freq, the only non-default parameter, was set low enough at $10^{-5}$ because the variant frequency can be as low as $10^{-3}$. Six replicates were all introduced for variant calling.

\paragraph{\textbf{VarScan2-somatic}}
We tested VarScan2-somatic on our six replicates synthetic data set as well. Considering that RVD2 doesn't call somatic status, we combined all the positions VarScan2-somatic called regardless the somatic status (Germline/LOH/Somatic/Unknown). --normal-purity was set at 1.00, while --tumor-purity was set as the dilution rate. --min-var-freq was set at $10^{-5}$.

\paragraph{\textbf{Strelka}}
Configuration and Analysis for Strelka is standard so we just installed the program and ran it on our data set. One thing to note is that since Strelka can not accept relicate data, we used a single replicate with a read depth that most closely matched the overall median depth of the replicates.

\paragraph{\textbf{muTect}}
Similar to Strelka, we ran muTect with standard settings and only passed one replicate of synthetic data.

GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples. Therefore, we expect those algorithms to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples. \citet{Stead:2013fu} showed that varscan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000. We find its performance on much lower MAF variants and across a wider range of coverage depths. Varscan2-somatic, strelka and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the overall median depth of the replicates.
